---
title: "LPJmL Runner"
author: "Jannes Breier"
date: "`r Sys.Date()`"
output:
  html_document: default
  # pdf_document: default
  md_document:
    variant: gfm
knit: (function(inputFile, encoding){
  rmarkdown::render(inputFile, encoding = encoding,
  output_format = "all") })
vignette: >
  %\VignetteIndexEntry{LPJmL Runner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**LPJmL Runner** `r if (!knitr::is_latex_output()) {"&#127939;"}`
is a lpjmlkit module of functions that have the goal to
simplify the execution of simulations with LPJmL and further to execute
complex, nested and multiple simulation sequences fast and less error prone
without having a big (bash) script overhead.  
To install LPJmL, read the [LPJmL installation instructions](https://github.com/PIK-LPJmL/LPJmL/blob/master/INSTALL).
\newline
\newline


## `r if (!knitr::is_latex_output()) {"&#9881;"}` Setup

**! Important !** The LPJmL Runner module only supports unix-based
operating systems in which the
[working environment for LPJmL](https://github.com/PIK-LPJmL/LPJmL/blob/master/INSTALL)
is configured!\
&nbsp;

## Overview

The LPJmL Runner generally requires 3 to 4 working steps:
Define a modified parameter table (1), create the corresponding configuration
files (2), check if the these are valid for LPJmL (3 - optional) and
run or submit LPJmL with each configurations (4).

#### **1. `r if (!knitr::is_latex_output()) {"&#128203;"}` Define a table of modified configuration parameters**\
  \
  Define what LPJmL configurations/parameters to be changed. Please familiarize
  yourself with available configuration options.  
  The base configuration file (e.g."lpjml_config.cjson") can be read in via
  `read_config` as a nested list object.
  Using the same syntax configurations/parameters can be changed directly in the
  corresponding configuration file or in a data frame (see example).  
  `?write_config` *for more information.*

  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  my_params <- tibble(
    sim_name = c("scenario1", "scenario2"),
    random_seed = c(42, 404),
    `param$k_temp` = c(NA, 0.03),
    new_phenology = c(TRUE, FALSE)
  )
  ```

\newpage

#### **2. `r if (!knitr::is_latex_output()) {"&#9997;"}` Create corresponding Configuration files**\
  \
  Now the central function is `write_config`, create and write
  LPJmL Configuration (config) file(s) `"config_*.json"` from a data frame
  with the parameters of a base config file `"lpjml_config.cjson"` to be
  changed.
  `?write_config` *for more information.*

  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  config_details <- write_config(my_params, model_path, sim_path)
  ```

#### **3. `r if (!knitr::is_latex_output()) {"&#128269;"}` Check validity of Configurations**\
  \
  Check whether your config(s) are valid for LPJmL by passing the returned
  data frame to `check_lpjml`. It won't raise an error (dependencies might not
  be satisfied yet) but will print/return the information of `lpjcheck`.

  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  lpjml_check(config_details, model_path, sim_path)
  ```

#### **4. `r if (!knitr::is_latex_output()) {"&#9654;"}` Run or `r if (!knitr::is_latex_output()) {"&#128640;"}` submit LPJmL**\
  \
  Run LPJmL for each Configuration locally via `run_lpjml` or submit as a 
  batch job to SLURM (PIK Cluster) via `submit_lpjml`. `run_lpjml` can also be
  utilized within slurm jobs to execute multiple single cell runs.\
  `?submit_lpjml` *or* `?run_lpjml` *for more information.*

  ```{r, eval = FALSE, highlight = TRUE}
  # run interactively
  run_details <- run_lpjml(config_details, model_path, sim_path)
  # OR submit to SLURM
  submit_details <- submit_lpjml(config_details, model_path, sim_path)
  ```

#### **miscellaneous** \
  \
  More helpful functions that come with LPJmL Runner are:

  * `read_config` to read a configuration file as a nested R list object

  * use the R internal `View` function for a tree visualization of a `"config_*.json"` file

  * `make_lpjml` function for compiling LPJmL.

&nbsp;
&nbsp;

## Usage

```{r setup, echo = TRUE, eval = FALSE, highlight = TRUE}
library(lpjmlkit)
# Why tibble? -> https://r4ds.had.co.nz/tibbles.html
# Tibbles also provide a better overview of the data and directly show the type
# of each column, which is very important for integer/floating point values.
library(tibble)

model_path <- "./LPJmL_internal"
sim_path <- "./my_runs"

```

&nbsp;
&nbsp;

\newpage

### Single cell simulations\
Single cell (or short number of multiple cells) simulations can be executed
locally or on a login node. This mode is especially useful when it comes to
testing or comparing local data.

#### **Example** *Potential natural vegetation and land-use run*  

&nbsp;

```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
# create parameter tibble
params <- tibble(
  sim_name = c("spinup", "lu", "pnv"),
  landuse = c("no", "yes", "no"),
  # only for demonstration
  nspinup = c(1000, NA, NA),
  reservoir = c(FALSE, TRUE, FALSE),
  startgrid = c(27410, 27410, 27410),
  river_routing = c(FALSE, FALSE, FALSE),
  wateruse = c("no", "yes", "no"),
  const_deposition = c(FALSE, FALSE, TRUE),
  # run parameter: dependency sets the restart paths to the corresponding
  #   restart_filename and calculates the execution order
  dependency = c(
    NA, "spinup", "spinup"
  )
)

# write config files
config_details <- write_config(
  x = params, # pass the defined parameter tibble
  model_path = model_path,
  sim_path = sim_path,
  js_filename = "lpjml_config.cjson" # (default) the base config file
)

# read and view config
config_lu <- read_config(
  filename = paste0(sim_path, "/configurations/config_lu.json") # nolint:absolute_path_linter.
)
View(config_lu)

# check config & LPJmL
check_config(
  x = config_details, # can be filename (vector) or tibble
  model_path = model_path,
  sim_path = sim_path
)

# execute runs sequentially
run_details <- run_lpjml(
  config_details,
  model_path = model_path,
  sim_path = sim_path
)

```


#### **Example** *Old vs. new phenology and old land-use vs. input toolbox*

&nbsp;

```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
# create parameter tibble
params <- tibble(
  sim_name = c("spinup_oldphen",
               "spinup_newphen",
               "oldphen",
               "old_lu",
               "lu_toolbox"),
  # object oriented like syntax to access nested json elements
  `input$landuse$name` = c(
    NA,
    NA,
    NA,
    NA,
    "input_toolbox_30arcmin/cftfrac_1500-2017_64bands_f2o.clm"
  ),
  nspinup = c(1000, 1000, NA, NA, NA),
  new_phenology = c(FALSE, TRUE, FALSE, TRUE, TRUE),
  startgrid = c(27410, 27410, 27410, 27410, 27410),
  river_routing = c(FALSE, FALSE, FALSE, FALSE, FALSE),
  dependency = c(NA, NA, "spinup_oldphen", "spinup_newphen", "spinup_newphen")
)

# write config files
config_details <- write_config(params, model_path, sim_path)

# check config & LPJmL
check_config(config_details, model_path, sim_path)

# execute runs sequentially
run_details <- run_lpjml(config_details, model_path, sim_path)

```

&nbsp;
&nbsp;
&nbsp;

\newpage
### Global simulations on the PIK cluster
Global simulations are simulations on all available cells with a coherent water
cycle. It requires more computational ressources which is why they have to be
run at dedicated compute nodes, at PIK Cluster only accessible via SLURM Job
scheduler. Therefore LPJmL has to be "submitted".

####  **Example** *Compare old vs new land use (lpjml input toolbox)*

&nbsp;

```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
# create parameter tibble
params <- tibble(
  sim_name = c("spinup",
               "old_lu",
               "lu_toolbox"),
  `input$landuse$name` = c(
    NA,
    NA,
    "input_toolbox_30arcmin/cftfrac_1500-2017_64bands_f2o.clm"
  ),
  dependency = c(NA, "spinup", "spinup"),
  # slurm option wtime: analogous to sbatch -wtime defines slurm option
  #   individually per config, overwrites submit_lpjml argument
  #   (same for sclass, ntasks, blocking or constraint)
  wtime = c("15:00:00", "3:00:00", "3:00:00")
)

# write config files
config_details <- write_config(
  x = params,
  model_path = model_path,
  sim_path = sim_path,
  output_list = c("vegc", "soilc", "cftfrac", "pft_harvestc", "irrig"),
  output_list_timestep = c("annual", "annual", "annual", "annual", "monthly"),
  output_format = "clm"
)

# check config & LPJmL
check_config(config_details, model_path, sim_path)

# submit runs to slurm
run_details <- submit_lpjml(
  x = config_details,
  model_path = model_path,
  sim_path = sim_path,
  group = "open"
)

```

&nbsp;
&nbsp;

\newpage

### Notes & tips

1. You can save the generated config tibble by applying `saveRDS` to it to reuse
for a rerun or resubmission next time ...

    ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
    saveRDS(config_details,
            paste0(sim_path, "/configurations/config_details.rds")) # nolint:absolute_path_linter.
    # next time ...
    config_details <- readRDS(
      paste0(sim_path, "/configurations/config_details.rds") # nolint:absolute_path_linter.
    )
    ```

2. Also if you want do not want to submit all runs you can ...

    ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
    # use a subset for the rows - in this example you may only want to resubmit the
    #   transient runs
    run_details <- submit_lpjml(
      x = config_details[2:3, ],
      model_path = model_path,
      sim_path = sim_path,
      group = "open"
    )
    ```

3. *a bit dirty though* If you want to reuse an old spinup simulation, you can copy the file or
create a symlink of the file to `"<sim_path>/restart/<spinup_sim_name>/restart.lpj"`.
Make sure the file/symlink is named `"restart.lpj"`
