---
title: "LPJmL Data"
author: "Jannes Breier"
date: "`r Sys.Date()`"
output:
  pdf_document
  # rmarkdown::html_vignette
  # md_document:
  #   variant: gfm
vignette: >
  %\VignetteIndexEntry{lpjml-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**LPJmL Data** is a lpjmlkit module that groups around the data class
`LPJmLData` and aims to facilitate the reading and processing of LPJmL inputs
and outputs by combining the raw data with available meta data (meta files,
header files or manually) to avoid a large overhead. It further enables easy
subsetting, transformations and basic statistics of the data and allows export
to common data formats.
All in all, with only a few lines of code.

&nbsp;

## Overview

LPJmL Data first requires the reading of LPJmL input or output data by applying
the read_io function (1). The returned object is of class LPJmLData (2) for
which base stats can be calculated (3), the inner data can be modified (4) or
common data formats can be exported (5).

#### **1. Data reading function read_io**\
  \
  The generic function to read LPJmL input and output, currently supported are
  three different file formats, "meta", "clm" and "raw":

* **"meta"** - *strongly recommended to use*.  
  For outputs from simulations in which `"output_metafile" : true` was 
  configured or inputs using the format.
  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  # read monthly runoff with meta file
  runoff <- read_i("./output/mrunoff.bin.json")
  ```

* **"clm"** - *use if "meta" is not available or in combination*.  
  For all common LPJmL inputs or for outputs from simulations in which
  `"fmt" : "raw"` was configured. Some meta information (e.g. `band_names`)
  have to be specified manually.
  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  # read monthly runoff data with header
  runoff <- read_i("./output/mrunoff.clm",
                   # set nbands and nstep manually otherwise month
                   #    dimension is not recognized
                   nbands = 1,
                   nstep = 12,
                   # not mandatory but useful information
                   variable = "runoff",
                   descr = "monthly runoff",
                   unit= "mm/month")
  ```

* **"raw"** - *not recommended to use (for LPJmL)*.  
  For old raw data. Every meta data has to be specified manually.
  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  # read monthly runoff raw data
  runoff <- read_i("./output/mrunoff.bin",
                   # every information has to be specified manually
                   ...)
  ```

&nbsp;

#### **2. Data class LPJmLData**\
  \
  read_io returns an object of a R6 class `LPJmLData` with two main attributes,
  `$data` and `$meta`:

* **\$data**  `class base::array` - returns the data array with default
  dimensions cell, time and band
  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  runoff$data
  #     , , band = 1
  # 
  #          time
  # cell       1901-01-31    1901-02-28    1901-03-31    1901-04-30
  #   0      2.427786e+02  1.265680e+02  2.279087e+02  2.027685e+02
  #   1      4.189225e-14  1.032507e-16  0.000000e+00  0.000000e+00
  #   2      3.860512e-14  0.000000e+00  0.000000e+00  0.000000e+00
  #   3      0.000000e+00  0.000000e+00  0.000000e+00  0.000000e+00
  #   4      0.000000e+00  0.000000e+00  0.000000e+00  0.000000e+00
  #   5      0.000000e+00  0.000000e+00  0.000000e+00  0.000000e+00
  ```
* **\$meta** Meta data of class `LPJmLMetaData` - returns the corresponding meta
  data (e.g. `runoff$meta$unit`)
  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  runoff$meta
  # $sim_name "lu_cf"
  # $source "LPJmL C Version 5.3.001"
  # $history "./LPJmL_internal/bin/lpjml ./configurations/config_lu_cf.json"
  # $variable "runoff"
  # $descr "monthly runoff"
  # $unit "mm/month"
  # $nbands 1
  # $nyear 111
  # $firstyear 1901
  # $lastyear 2011
  # $nstep 12
  # $timestep 1
  # $ncell 67420
  # $firstcell 0
  # $cellsize_lon 0.5
  # $cellsize_lat 0.5
  # $datatype "float"
  # $scalar 1
  # $order "cellseq"
  # $bigendian FALSE
  # $format "raw"
  # $filename "runoff.bin"
  ```
&nbsp;

#### **3. Base stats of LPJmLData objects**\
  \
  To get an overview of the data, `LPJmLData` offers support
  for various base functions: `length()`, `dim()`, `dimension()`, `summary()` and
  `plot()`.  
  *More methods can be added uppon request.*

```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
# self print, also via print(runoff)
runoff
# $meta %>%
#   .$sim_name "lu_cf"
#   .$variable "runoff"
#   .$descr "monthly runoff"
#   .$unit "mm/month"
#   .$nbands 1
#   .$nyear 111
#   .$nstep 12
#   .$timestep 1
#   .$ncell 67420
#   .$cellsize_lon 0.5
#   .$cellsize_lat 0.5
# Note: not printing all meta data, use $meta to get all.
# $data %>%
#   dimnames() %>%
#     .$cell  "0" "1" "2" "3" ... "67419"
#     .$time  ""1901-01-31" "1901-02-28" "1901-03-31" "1901-04-30" ... "2011-12-31"
#     .$band  "1"
# $summary()
#        1
#  Min.   :   0.0000
#  1st Qu.:   0.0619
#  Median :   4.4320
#  Mean   :  28.7658
#  3rd Qu.:  27.5627
#  Max.   :2840.9602
# Note: summary is not weighted by grid area.

# dimension length of $data array, dimnames function also available
dim(runoff)
#  cell  time  band
# 67420  1332     1

# plot as maps or time series, depending on the dimensions
plot(runoff)
```

&nbsp;

#### **4. Modify LPJmLData objects**\
  \
  LPJmLData objects come with a bundle of methods to modify its state: 
  `add_grid()`, `transform()` and `subset()`.

* **`add_grid()`** Add a **$grid** attribute (as LPJmLData object) to the
  object, providing the spatial reference (longitude and latitude) for every
  cell.
  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  # object oriented (R6 class) notation (assigning grid directly to runoff)
  runoff$add_grid()

  # common R notation (overwriting the original object)
  runoff <- add_grid(runoff)

  # use read_io arguments if file_type != meta
  runoff <- add_grid(runoff, "./output/grid.clm")
  ```

* **`transform()`** Transform the `$data` dimensions.  
  Either the cell dimension into two "lon" (longitude) and "lat" (latitude) 
  dimensions or the time into "year", "month" and "day" dimension 
  (if available) - combinations and back transformations are also possible. 
  For format "lon_lat" attribute `$grid` is required.
  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  # transform into lon and lat dimensions, if add_grid is not executed
  #   before, it is called implicitly, if file_format == "meta"
  runoff <- transform(runoff, to = "lon_lat")
  runoff
  # [...]
  # $data %>%
  #   dimnames() %>%
  #     .$lat  "-55.75" "-55.25" "-54.75" "-54.25" ... "83.75"
  #     .$lon  "-179.75" "-179.25" "-178.75" "-178.25" ... "179.75"
  #     .$time  "1901-01-31" "1901-02-28" "1901-03-31" "1901-04-30" ... "2011-12-31"
  #     .$band  "1"
  # [...]

  # transform into year and month dimension (day not available for montly
  #   runoff)
  runoff <- transform(runoff, to = "year_month_day")
  runoff
  # [...]
  # $data %>%
  #   dimnames() %>%
  #     .$lat  "-55.75" "-55.25" "-54.75" "-54.25" ... "83.75"
  #     .$lon  "-179.75" "-179.25" "-178.75" "-178.25" ... "179.75"
  #     .$month  "1" "2" "3" "4" ... "12"
  #     .$year  "1901" "1902" "1903" "1904" ... "2011"
  #     .$band  "1"
  # [...]

  # transform back to original dimensions
  runoff <- transform(runoff, to = c("cell", "time"))
  runoff
  # [...]
  # $data %>%
  #   dimnames() %>%
  #     .$cell  "0" "1" "2" "3" ... "67419"
  #     .$time  "1901-01-31" "1901-02-28" "1901-03-31" "1901-04-30" ... "2100-12-31"
  #     .$band  "1"
  # [...]
  ```

* **`subset()`** Subset the `$data`.  
  Use `$data` dimensions as key and dimension names or indices as value to
  subset `$data`. `$meta` is adjusted according to the subset.
  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  # subset by dimnames (character string)
  runoff <- subset(runoff, cell = "1991-05-31")
  runoff
  # $meta %>%
  #   .$nyear 1
  #   .$ncell 67420
  #   .$subset TRUE
  # [...]
  # Note: not printing all meta data, use $meta to get all.
  # $data %>%
  #   dimnames() %>%
  #     .$cell  "0" "1" "2" "3" ... "67419"
  #     .$time  "1991-05-31"
  #     .$band  "1"
  # [...]

  # subset by indices
  runoff <- subset(runoff, cell = 28697:28700)
  runoff
  # $meta %>%
  #   .$nyear 1
  #   .$ncell 4
  #   .$subset TRUE
  # [...]
  # Note: not printing all meta data, use $meta to get all.
  # $data %>%
  #   dimnames() %>%
  #     .$cell  "28696" "28697" "28698" "28699"
  #     .$time  "1991-05-31"
  #     .$band  "1"
  # [...]
  ```
&nbsp;

#### **5. Export LPJmLData objects**\
  \
  Finally LPJmLData objects can be exported into common R data
  formats: `array`, `tibble`, `raster` and `terra`.  
  *More export methods can be added uppon request.*

* **`as_array()`** Export as `array`. Besides directly accessing `$data`
  as_array further provides functionality to subset and aggregate `$data`.
  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  # export as array with subset of first 4 time steps and aggregation of
  #   dimension cell (mean)
  as_array(runoff,
           subset = list(time = 1:6),
           aggregate = list(cell = mean))
  #             band
  # time                1
  #   1901-01-31 19.49611
  #   1901-02-28 20.28368
  #   1901-03-31 27.93595
  #   1901-04-30 36.90505
  #   1901-05-31 39.38885
  #   1901-06-30 32.80252
  ```

* **`as_tibble()`** Export `$data` as a
  [`tibble`](https://tibble.tidyverse.org/) object, providing the same further
  functionality as as_array.
  as_array further provides functionality to subset and aggregate `$data`.
  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  # export as array with subset of first 4 time steps and aggregation of
  #   dimension cell (mean)
  as_tibble(runoff, subset = list(time = 1:6))
  # # A tibble: 404,520 × 4
  #    cell  time       band  value
  #    <fct> <fct>      <fct> <dbl>
  #  1 0     1901-01-31 1      184.
  #  2 1     1901-01-31 1        0 
  #  3 2     1901-01-31 1        0 
  #  4 3     1901-01-31 1        0 
  #  5 4     1901-01-31 1        0 
  #  6 5     1901-01-31 1        0 
  #  7 6     1901-01-31 1        0 
  #  8 7     1901-01-31 1        0 
  #  9 8     1901-01-31 1        0 
  # 10 9     1901-01-31 1        0 
  # # … with 404,510 more rows
  ```

* **`as_raster()`** / **`as_terra()`** ** Export `$data` as a
  [`raster`](https://rspatial.github.io/raster/reference/raster-package.html)
  or a [`terra`](https://rspatial.org/) object (successor of
  raster), providing the same further functionality as as_array.
  ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
  # export first time step as raster
  as_raster(runoff, subset = list(time = 1))
  # class      : RasterLayer 
  # dimensions : 280, 720, 201600  (nrow, ncol, ncell)
  # resolution : 0.5, 0.5  (x, y)
  # extent     : -180, 180, -56, 84  (xmin, xmax, ymin, ymax)
  # crs        : +proj=longlat +datum=WGS84 +no_defs 
  # source     : memory
  # names      : runoff 
  # values     : -1.682581e-13, 671.8747  (min, max)

  # export first time step as terra SpatRaster
  as_terra(runoff, subset = list(time = 1))
  # class       : SpatRaster 
  # dimensions  : 280, 720, 1  (nrow, ncol, nlyr)
  # resolution  : 0.5, 0.5  (x, y)
  # extent      : -180, 180, -56, 84  (xmin, xmax, ymin, ymax)
  # coord. ref. : lon/lat WGS 84 (EPSG:4326) 
  # source      : memory 
  # name        :        runoff 
  # min value   : -1.682581e-13 
  # max value   :  6.718747e+02 
  # unit        :      mm/month

  # export first 4 time step as raster brick
  as_raster(runoff, subset = list(time = 1:4))
  # class      : RasterBrick 
  # dimensions : 280, 720, 201600, 4  (nrow, ncol, ncell, nlayers)
  # resolution : 0.5, 0.5  (x, y)
  # extent     : -180, 180, -56, 84  (xmin, xmax, ymin, ymax)
  # crs        : +proj=longlat +datum=WGS84 +no_defs 
  # source     : memory
  # names      :   X1901.01.31,   X1901.02.28,   X1901.03.31,   X1901.04.30
  # min values : -1.682581e-13, -1.750495e-13, -2.918900e-13, -1.516298e-13
  # max values :      671.8747,      785.2363,      828.2853,      987.4359

  # export first 4 time step as terra SpatRaster
  as_terra(runoff, subset = list(time = 1:4))
  # class       : SpatRaster 
  # dimensions  : 280, 720, 4  (nrow, ncol, nlyr)
  # resolution  : 0.5, 0.5  (x, y)
  # extent      : -180, 180, -56, 84  (xmin, xmax, ymin, ymax)
  # coord. ref. : lon/lat WGS 84 (EPSG:4326) 
  # source      : memory 
  # names       :    1901-01-31,    1901-02-28,    1901-03-31,    1901-04-30 
  # min values  : -1.682581e-13, -1.750495e-13, -2.918900e-13, -1.516298e-13 
  # max values  :  6.718747e+02,  7.852363e+02,  8.282853e+02,  9.874359e+02 
  # unit        :      mm/month,      mm/month,      mm/month,      mm/month 
  # time (days) : 1901-01-31 to 1901-04-30 
  ```
&nbsp;

#### **miscellaneous** \
  \
  More helpful functions that come with LPJmL Data are:

  * `read_meta` to read meta meta information of meta and header files as 
    `LPJmLMetaData` objects

  * `LPJmLMetaData` objects can be exported as `as_list` and `as_header` to use
    for creating header objects or write header files.


&nbsp;
&nbsp;

## Usage

```{r setup, echo = TRUE, eval = FALSE, highlight = TRUE}
library(lpjmlkit)
```

#### **1. Example** *Global Trend in  npp over the years*

&nbsp;

```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
npp <- read_io(filename = "./output/npp.bin.json",
               subset = list(year = 1970:2011))

# transform time in year and month dimension
npp$transform(to = "year_month_day")

# plot timeseries with aggregated cell and month dimension
plot(npp,
     aggregate = list(cell = sum, month = sum))

# also available as data array
global_npp_trend <- as_array(npp,
                             aggregate = list(cell = sum, month = sum))


```


#### **2. Example** *Runoff in northern hemisphere's summertime*

&nbsp;

```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
# alternative function notation via %>% operator (in magrittr package)
library("magrittr")
runoff <- read_io(filename = "./output/runoff.bin.json",
               subset = list(year = 2002:2011))

runoff %>%
  # transform time and space dimensions
  transform(to = c("year_month_day", "lon_lat")) %>%
  # ... to subset summer month as well as northern hemisphere (positive) 
  #   latitudes
  subset(month = 6:9,
         lat = seq(0.25, 83.75, by = 0.5)) %>%
  # for plotting sum up summer month and take the average over the years
  plot(aggregate = list(year = mean, month = sum))

```


#### **3. Example** *GPP per latitude*

&nbsp;

```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
# alternative function notation via %>% operator (in magrittr package)
gpp <- read_io(filename = "./output/gpp.bin.json",
               subset = list(year = 2002:2011))

# transform into lon_lat format
gpp$transform(to = "lon_lat")

# plot gpp per latitude
plot(gpp, aggregate = list(time = mean, lon = mean))

```


#### **4. Example** *CFT fractions for area around Potsdam*

&nbsp;

```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
# coordinates for cells around Potsdam
coordinates <- tibble::tibble(lat = c(52.25, 52.400922, 53.25),
                      lon = c(12.75, 13.03638, 12.75))

# complete pipe notation, from reading to plotting data
read_io(
  filename = glue("{output_path}/cftfrac.bin.json"),
  subset = list(year = 2000:2018)
) %>%
  transform(to = "lon_lat") %>%
  # special case for subsetting of lat and lon pairs
  subset(coords = coordinates) %>%
  # mean spatial dimensions
  plot(aggregate = list(lon = mean, lat = mean))

```

&nbsp;
&nbsp;

### Notes & tips

1. `LPJmLData` and `LPJmLMetaData` objects are closed environments, each of an
  R6 class that function as a data container.  
  Do not replicate R6 objects like  
    ```{r, echo = TRUE, eval = FALSE, highlight = TRUE}
      my_data <- lpjml_data
      # instead use
      my_data <- lpjml_data$clone(deep = TRUE)
    ```
  Otherwise if any methods are performed on `my_data` the same method is
  performed on `lpjml_data` since they point to the same environment.

2. Do not try to overwrite any `LPJmLData` objects. It is either not
  possible or can mess up the integrity of the object. Therefore do not use
  method named like `$.__<method>__` or attributes named like
  `$._<attribute>_`

3. When performance is important, choose R6 method notation
  `runoff$transform(to = "lon_lat")` over common R notation
  `transform(runoff, to = "lon_lat")`.

4. When comparing old (< LPJmL version 5.3) data with LPJmL 5.3 data it can
  be useful to combine meta (`"output_metafile" : true`) with the header file
  format (`"fmt": "clm"`) for simplification of process pipelines.

